Lessons learned and paths forward for rabies dog vaccination in Madagascar: a case study of pilot vaccination campaigns in Moramanga District

### 2021-04-05 paper/mdpi_paper.Rmd
post-acceptance files

  - name: Annie Yang

@@

   We want to particularly acknowledge the contributions to this work by Annie Yang, 

@@

  Conceptualization, MR, GTE, MA, JN, ZF, TS, JL and KV; Data curation, CF, MR, ZR, CH, RR, AF, AY, FD, JL and TR; Formal      analysis, CF, MR, AY and CJEM; Funding acquisition, MR, ZF, TS, JL, CJEM and KV; Investigation, MR, ZR, CH, RR, AF, TAC and   TR; Methodology, MR, ZR, CH, GTE, MA, AF, AY, ZF, JL, TR, KH, CJEM and KV; Project administration, ZR, CH, RR, GTE, MA,      NPR, JN, AF, FD, TAC, ZF, TS, JL, TR and KV; Supervision, GTE, MA, NPR, JN, ZF, KH, CJEM and KV; Visualization, MR;          Writing---original draft, CF and KV; Writing---review & editing, CF, MR, ZR, CH, GTE, TAC, ZF, KH, CJEM and KV.

@@

In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga (fig 2). This campaign focused on two communes in Moramanga, Moramanga Ville (the district center) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of probable rabies exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].

The campaign was planned as a series of static point vaccination stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided fliers advertising the date of the vaccine and that it would be available at no cost to owners (fig 2). During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old based on current WHO recommendations for endemic settings [@morters2015; @world2018expert]. Rabisin has a manufacturor stated duration of protection of 1 year given one dose, and of an additional three years if an additional dose is given approximately one year after the first dose. As part of the campaign, owners were surveyed by vaccinators about how many dogs and cats they owned in total (split by > 1 year vs. < 1 year in order to avoid language ambiguities that might result in excluding puppies and kittens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mirenyreny', tied/'mifatora', or fenced/'mifefy'). Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, owners were asked how much in Ariary they would be willing to pay to have one animal vaccinated against rabies (after being informed that the current vaccination was being free). For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years as reported by the owner, and whether the animal had been previously vaccinated.

@@

Through vaccination campaigns implemented in Moramanga District of Madagascar, we saw high demand for vaccination from dog owners and found that dogs were accessible and able to be handled safely and efficiently for parenteral vaccination at a reasonable cost (between 1.3 - 1.8 USD per animal vaccinated). We find that providing the vaccine at no direct cost to dog owners will be critical to achieving sufficient coverage, as even with nominal fees, a significant proportion of owners indicated they would no longer vaccinate their animals. A static point vaccination strategy achieved higher coverage over a shorter time period in 2018 compared to dog vaccinations conducted by the District Veterinarian as part of routine veterinary service provision in 2019. However, it came at a higher cost per animal vaccinated, was more limited in geographic scope, and required more resources in terms of personnel. In addition, in the rural setting of Andasibe, the static point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location. Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners (fig 2), we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.

>>>>

  - name: Annie L. Yang

@@

   We want to particularly acknowledge the contributions to this work by Annie Li Yang, 

@@

  Conceptualization, MR, GTE, MA, JN, ZF, TS, JL and KV; Data curation, CF, MR, ZR, CH, RR, AF, ALY, FD, JL and TR; Formal      analysis, CF, MR, ALY and CJEM; Funding acquisition, MR, ZF, TS, JL, CJEM and KV; Investigation, MR, ZR, CH, RR, AF, TAC and   TR; Methodology, MR, ZR, CH, GTE, MA, AF, ALY, ZF, JL, TR, KH, CJEM and KV; Project administration, ZR, CH, RR, GTE, MA,      NPR, JN, AF, FD, TAC, ZF, TS, JL, TR and KV; Supervision, GTE, MA, NPR, JN, ZF, KH, CJEM and KV; Visualization, MR; Writing---original draft, CF and KV; Writing---review & editing, CF, MR, ZR, CH, GTE, TAC, ZF, KH, CJEM and KV.

@@

In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga (fig 1). This campaign focused on two communes in Moramanga, Moramanga Ville (the district center) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of probable rabies exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].

The campaign was planned as a series of static point vaccination stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided fliers advertising the date of the vaccine and that it would be available at no cost to owners (fig 1). During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old based on current WHO recommendations for endemic settings [@morters2015; @world2018expert]. Rabisin has a manufacturor stated duration of protection of 1 year given one dose, and of an additional three years if an additional dose is given approximately one year after the first dose. As part of the campaign, owners were surveyed by vaccinators about how many dogs and cats they owned in total (split by > 1 year vs. < 1 year in order to avoid language ambiguities that might result in excluding puppies and kittens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mirenyreny', tied/'mifatora', or fenced/'mifefy'). Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, owners were asked how much in Ariary they would be willing to pay to have one animal vaccinated against rabies (after being informed that the current vaccination was being free). For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years as reported by the owner, and whether the animal had been previously vaccinated.

@@

Through vaccination campaigns implemented in Moramanga District of Madagascar, we saw high demand for vaccination from dog owners and found that dogs were accessible and able to be handled safely and efficiently for parenteral vaccination at a reasonable cost (between 1.3 - 1.8 USD per animal vaccinated). We find that providing the vaccine at no direct cost to dog owners will be critical to achieving sufficient coverage, as even with nominal fees, a significant proportion of owners indicated they would no longer vaccinate their animals. A static point vaccination strategy achieved higher coverage over a shorter time period in 2018 compared to dog vaccinations conducted by the District Veterinarian as part of routine veterinary service provision in 2019. However, it came at a higher cost per animal vaccinated, was more limited in geographic scope, and required more resources in terms of personnel. In addition, in the rural setting of Andasibe, the static point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location. Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners (see puppies picutred in basket in fig 1), we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.

### 2021-04-05 paper/mdpi_paper.tex
post-acceptance files

In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal
Doctors (TAD)) organized a pilot vaccination campaign in collaboration
with the Department of Veterinary Services, and the Ministry of Public
Health in the District of Moramanga (fig 2). This campaign focused on
two communes in Moramanga, Moramanga Ville (the district center) and
Andasibe (a rural commune surrounding Andasibe National Park), where
previously high incidence of probable rabies exposures (Moramanga Ville)
and a high burden of deaths (Andasibe) had been recorded
\citep{rajeev2018}.

The campaign was planned as a series of static point vaccination
stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week
before the campaign dates, the vaccination team informed the chief of
the fokontany about the campaign and provided fliers advertising the
date of the vaccine and that it would be available at no cost to owners
(fig 2). During the campaign, we used Rabisin (10 mL vials with 1 mL per
dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented
that were at least 1 month old based on current WHO recommendations for
endemic settings \citep{morters2015, world2018expert}. Rabisin has a
manufacturor stated duration of protection of 1 year given one dose, and
of an additional three years if an additional dose is given
approximately one year after the first dose. As part of the campaign,
owners were surveyed by vaccinators about how many dogs and cats they
owned in total (split by \textgreater{} 1 year vs.~\textless{} 1 year in
order to avoid language ambiguities that might result in excluding
puppies and kittens), as well as if their dogs were free roaming (no
restrictions on movement by the owner/`mirenyreny', tied/`mifatora', or
fenced/`mifefy'). Vaccinations were delivered at no cost to owners, but
as animal vaccination is generally thought of as a paid service in
Madagascar, owners were asked how much in Ariary they would be willing
to pay to have one animal vaccinated against rabies (after being
informed that the current vaccination was being free). For each animal
vaccinated, we recorded the species (cat or dog), sex, approximate age
in years as reported by the owner, and whether the animal had been
previously vaccinated.

>>>>

In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal
Doctors (TAD)) organized a pilot vaccination campaign in collaboration
with the Department of Veterinary Services, and the Ministry of Public
Health in the District of Moramanga (fig 1). This campaign focused on
two communes in Moramanga, Moramanga Ville (the district center) and
Andasibe (a rural commune surrounding Andasibe National Park), where
previously high incidence of probable rabies exposures (Moramanga Ville)
and a high burden of deaths (Andasibe) had been recorded
\citep{rajeev2018}.

The campaign was planned as a series of static point vaccination
stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week
before the campaign dates, the vaccination team informed the chief of
the fokontany about the campaign and provided fliers advertising the
date of the vaccine and that it would be available at no cost to owners
(fig 1). During the campaign, we used Rabisin (10 mL vials with 1 mL per
dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented
that were at least 1 month old based on current WHO recommendations for
endemic settings \citep{morters2015, world2018expert}. Rabisin has a
manufacturor stated duration of protection of 1 year given one dose, and
of an additional three years if an additional dose is given
approximately one year after the first dose. As part of the campaign,
owners were surveyed by vaccinators about how many dogs and cats they
owned in total (split by \textgreater{} 1 year vs.~\textless{} 1 year in
order to avoid language ambiguities that might result in excluding
puppies and kittens), as well as if their dogs were free roaming (no
restrictions on movement by the owner/`mirenyreny', tied/`mifatora', or
fenced/`mifefy'). Vaccinations were delivered at no cost to owners, but
as animal vaccination is generally thought of as a paid service in
Madagascar, owners were asked how much in Ariary they would be willing
to pay to have one animal vaccinated against rabies (after being
informed that the current vaccination was being free). For each animal
vaccinated, we recorded the species (cat or dog), sex, approximate age
in years as reported by the owner, and whether the animal had been
previously vaccinated.
